---
title: "✨ Loops and Conditionals in R 💻"
format: html
editor: visual
autho: Dr Chirag Parsania
---

# 🚀 Introduction

In this tutorial, we'll explore how to use `for` loops and `if` statements in R. These are essential tools for controlling the flow of your code.

------------------------------------------------------------------------

# 🔁 1. `for` Loops with Vectors

A `for` loop allows you to repeat a block of code for each element in a sequence.

## 🧪 Example: Loop over a vector

```{r}
# Define a vector of numbers
numbers <- c(1, 2, 3, 4, 5)

# Loop through each number and print it
for (n in numbers) {
  print(n)
}
```

------------------------------------------------------------------------

# 🧠 2. `if` Statements with Vectors

The `if` statement is used to perform conditional execution.

## 🔍 Example: Check for even or odd

```{r}
for (n in numbers) {
  if (n %% 2 == 0) {
    print(paste(n, "is even ✅"))
  } else {
    print(paste(n, "is odd 🔹"))
  }
}
```

------------------------------------------------------------------------

# 🗃️ 3. Looping Over Data Frames

You can loop through rows or columns of a data frame.

## 📋 Example: Loop through rows

```{r}
# Sample data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 17)
)

# Loop through each row and check if the person is an adult
for (i in 1:nrow(df)) {
  if (df$age[i] >= 18) {
    print(paste(df$name[i], "is an adult 👨‍💼"))
  } else {
    print(paste(df$name[i], "is a minor 🧒"))
  }
}
```

------------------------------------------------------------------------

# 🧺 4. Looping Over Lists

Lists are very flexible and can contain different types of elements.

## 📦 Example: Loop through a list

```{r}
# Define a list
my_list <- list(
  vec = c(1, 2, 3),
  df = data.frame(a = 1:2, b = 3:4),
  msg = "Hello"
)

# Loop and print type of each element
for (item in my_list) {
  print(class(item))
}
```

## 🧮 `if` with Lists

```{r}
for (item in my_list) {
  if (is.data.frame(item)) {
    print("📊 This is a data frame")
  } else if (is.vector(item)) {
    print("📈 This is a vector")
  } else {
    print("🔧 Other type")
  }
}
```

------------------------------------------------------------------------

# 📝 Summary

✅ Use `for` loops to iterate over vectors, data frames, or lists.\
✅ Use `if`, `else if`, and `else` for conditional logic.\
✅ Combine them for powerful control over your data workflows!

------------------------------------------------------------------------

Happy coding! ✨🧬💻
