---
title: "Cool Inbuilt R Functions"
author: "CI Bioinformatics Club"
format: html
editor: visual
---

# 📚 What You'll Practice

1.  [🔢 1. mean(): Calculate the Mean](#mean-calculate-the-mean)
2.  [📊 2. median(): Calculate the Median](#median-calculate-the-median)
3.  [🔀 3. mode(): Get the Mode (most frequent value)](#mode-get-the-mode-most-frequent-value)
4.  [🎲 4. sample(): Sample Random Elements](#sample-sample-random-elements)
5.  [🏆 5. sort(): Sort Data in Ascending or Descending Order](#sort-sort-data-in-ascending-or-descending-order)
6.  [↕️ 6. order(): Get Order of Elements]
7.  [🔎 7. unique(): Get Unique Values](#unique-get-unique-values)
8.  [♻️ 8. rev(): Reverse a Vector](#rev-reverse-a-vector)
9.  [🔄 9. rep(): Repeat Elements](#rep-repeat-elements)
10. [📊 10. summary(): Get Summary Statistics](#summary-get-summary-statistics)
11. [📅 11. seq(): Generate Sequences](#seq-generate-sequences)
12. [📁 12. str(): Structure of Object](#str-structure-of-object)
13. [📈 13. head() and tail(): Peek at Data](#head-and-tail-peek-at-data)
14. [📊 14. table(): Frequency Table](#table-frequency-table)
15. [➕ 15. paste() and paste0(): Concatenate Strings]
16. [16. cut() Divide Numeric Data into Groups]
17. [17. round(), floor(), ceiling(): Rounding Functions]
18. [📐 18. which(): Get Index of Matching Condition]
19. [💯 19. any() and all(): Logical Check Functions]
20. [🧪 20. is.na(): Check for Missing Values]
21. [📉 21. na.omit(): Remove Missing Values]
22. [🧬 22. length(): Number of Elements]
23. [🧩 23. dim(): Dimensions of Matrix or Data Frame]
24. [24. names(): Access or Set Names of Elements]
25. [📦 25. append(): Add Elements to a Vector]

# 🧠 Introduction

In this tutorial, we will explore some of the most commonly used and cool **inbuilt R functions**. These functions are great for performing simple tasks like calculating averages, sorting data, or even working with sample data. Let's dive in!

------------------------------------------------------------------------

# 🔢 1. mean(): Calculate the Mean {#mean-calculate-the-mean}

```{r}
# Calculate the mean of a numeric vector
numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
mean_value <- mean(numbers)
mean_value
```

------------------------------------------------------------------------

# 📊 2. median(): Calculate the Median {#median-calculate-the-median}

```{r}
# Calculate the median of a numeric vector
median_value <- median(numbers)
median_value
```

------------------------------------------------------------------------

# 🔀 3. mode(): Get the Mode (most frequent value) {#mode-get-the-mode-most-frequent-value}

```{r}
# Mode of a numeric vector (R doesn't have a built-in mode function, so we use a custom one)
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

mode_value <- getmode(c(1,2,2,3,4,4,4,5,5))
mode_value
```

------------------------------------------------------------------------

# 🎲 4. sample(): Sample Random Elements {#sample-sample-random-elements}

```{r}
# Sample 3 random elements from a vector
random_sample <- sample(numbers, 3)
random_sample
```

------------------------------------------------------------------------

# 🏆 5. sort(): Sort Data in Ascending or Descending Order {#sort-sort-data-in-ascending-or-descending-order}

```{r}
# Sort the numbers in ascending order
sorted_asc <- sort(numbers)
sorted_asc

# Sort the numbers in descending order
sorted_desc <- sort(numbers, decreasing = TRUE)
sorted_desc
```

------------------------------------------------------------------------

# ↕️ 6. order(): Get Order of Elements

```{r}
# Get the order of elements in a vector
values <- c(20, 10, 40, 30)
order(values)         # Indices that would sort the vector
values[order(values)] # Sorted values using order
```

------------------------------------------------------------------------

# 🔎 7. unique(): Get Unique Values {#unique-get-unique-values}

```{r}
# Get unique values from a vector
repeats <- c(1, 2, 2, 3, 3, 3, 4)
unique(repeats)
```

------------------------------------------------------------------------

# ♻️ 8. rev(): Reverse a Vector {#rev-reverse-a-vector}

```{r}
# Reverse the elements of a vector
rev(numbers)
```

------------------------------------------------------------------------

# 🔄 9. rep(): Repeat Elements {#rep-repeat-elements}

```{r}
# Repeat each element 3 times
rep(c("A", "B"), each = 3)

# Repeat entire vector 2 times
rep(c("A", "B"), times = 2)
```

------------------------------------------------------------------------

# 📊 10. summary(): Get Summary Statistics {#summary-get-summary-statistics}

```{r}
# Get summary statistics of a numeric vector
summary(numbers)
```

------------------------------------------------------------------------

# 📅 11. seq(): Generate Sequences {#seq-generate-sequences}

```{r}
# Generate a sequence from 1 to 10 by 2
seq(1, 10, by = 2)
```

------------------------------------------------------------------------

# 📁 12. str(): Structure of Object {#str-structure-of-object}

```{r}
# Check the structure of a data frame
example_df <- data.frame(Name = c("Tom", "Sara"), Age = c(25, 28))
str(example_df)
```

------------------------------------------------------------------------

# 📈 13. head() and tail(): Peek at Data {#head-and-tail-peek-at-data}

```{r}
# First few elements
head(numbers, 3)

# Last few elements
tail(numbers, 3)
```

------------------------------------------------------------------------

# 📊 14. table(): Frequency Table {#table-frequency-table}

```{r}
# Frequency table of values
table(c("apple", "banana", "apple", "cherry"))
```

------------------------------------------------------------------------

# ➕ 15. paste() and paste0(): Concatenate Strings

```{r}
# Paste with separator
paste("Hello", "world", sep = " ")

# Paste without separator
paste0("gene", 1:3)
```

------------------------------------------------------------------------

# 16. cut() Divide Numeric Data into Groups

```{r}
# Divide values into 3 groups
x <- c(1, 5, 10, 15, 20)
cut(x, breaks = 3)
```

------------------------------------------------------------------------

# 17. round(), floor(), ceiling(): Rounding Functions

```{r}
val <- 3.75
round(val)    # Round to nearest
floor(val)    # Round down
ceiling(val)  # Round up
```

------------------------------------------------------------------------

# 📐 18. which(): Get Index of Matching Condition

```{r}
# Index of elements > 5
which(numbers > 5)
```

------------------------------------------------------------------------

# 💯 19. any() and all(): Logical Check Functions

```{r}
# Check if any or all elements match a condition
any(numbers > 5)  # TRUE if any value is > 5
all(numbers > 5)  # TRUE only if all values are > 5
```

------------------------------------------------------------------------

# 🧪 20. is.na(): Check for Missing Values

```{r}
data <- c(1, 2, NA, 4)
is.na(data)       # Logical vector indicating NA
sum(is.na(data))  # Count of NAs
```

------------------------------------------------------------------------

# 📉 21. na.omit(): Remove Missing Values

```{r}
# Remove missing values from a vector
data_no_na <- na.omit(data)
data_no_na
```

------------------------------------------------------------------------

# 🧬 22. length(): Number of Elements

```{r}
# Get the number of elements in a vector
length(numbers)
```

------------------------------------------------------------------------

# 🧩 23. dim(): Dimensions of Matrix or Data Frame

```{r}
m <- matrix(1:12, nrow = 3)
dim(m)
```

------------------------------------------------------------------------

# 24. names(): Access or Set Names of Elements

```{r}
named_vec <- c(a = 1, b = 2, c = 3)
names(named_vec)
names(named_vec) <- c("x", "y", "z")
named_vec
```

------------------------------------------------------------------------

# 📦 25. append(): Add Elements to a Vector

```{r}
vec <- c(1, 2, 3)
append(vec, 4)           # Add 4 at the end
append(vec, 0, after = 0) # Add 0 at the beginning
```

------------------------------------------------------------------------

# ✅ Conclusion

These built-in R functions are essential tools for data manipulation and exploration. Get comfortable with them, and you'll have a strong foundation for more advanced work!
