---
title: "Cool Inbuilt R Functions"
author: "CI Bioinformatics Club"
format: html
editor: visual
---

# 🧠 Introduction

In this tutorial, we will explore some of the most commonly used and cool **inbuilt R functions**. These functions are great for performing simple tasks like calculating averages, sorting data, or even working with sample data. Let's dive in!

---

# 🔢 1. `mean()`: Calculate the Mean

```{r}
# Calculate the mean of a numeric vector
numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
mean_value <- mean(numbers)
mean_value
```

---

# 📊 2. `median()`: Calculate the Median

```{r}
# Calculate the median of a numeric vector
median_value <- median(numbers)
median_value
```

---

# 🔀 3. `mode()`: Get the Mode (most frequent value)

```{r}
# Mode of a numeric vector (R doesn't have a built-in mode function, so we use a custom one)
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

mode_value <- getmode(c(1,2,2,3,4,4,4,5,5))
mode_value
```

---

# 🎲 4. `sample()`: Sample Random Elements

```{r}
# Sample 3 random elements from a vector
random_sample <- sample(numbers, 3)
random_sample
```

---

# 🏆 5. `sort()`: Sort Data in Ascending or Descending Order

```{r}
# Sort the numbers in ascending order
sorted_asc <- sort(numbers)
sorted_asc

# Sort the numbers in descending order
sorted_desc <- sort(numbers, decreasing = TRUE)
sorted_desc
```

---

# ↕️ 6. `order()`: Get Order of Elements

```{r}
# Get the order of elements in a vector
values <- c(20, 10, 40, 30)
order(values)         # Indices that would sort the vector
values[order(values)] # Sorted values using order
```

---

# 🔎 7. `unique()`: Get Unique Values

```{r}
# Get unique values from a vector
repeats <- c(1, 2, 2, 3, 3, 3, 4)
unique(repeats)
```

---

# ♻️ 8. `rev()`: Reverse a Vector

```{r}
# Reverse the elements of a vector
rev(numbers)
```

---

# 🔄 9. `rep()`: Repeat Elements

```{r}
# Repeat each element 3 times
rep(c("A", "B"), each = 3)

# Repeat entire vector 2 times
rep(c("A", "B"), times = 2)
```

---

# 📊 10. `summary()`: Get Summary Statistics

```{r}
# Get summary statistics of a numeric vector
summary(numbers)
```

---

# 📅 11. `seq()`: Generate Sequences

```{r}
# Generate a sequence from 1 to 10 by 2
seq(1, 10, by = 2)
```

---

# 📁 12. `str()`: Structure of Object

```{r}
# Check the structure of a data frame
example_df <- data.frame(Name = c("Tom", "Sara"), Age = c(25, 28))
str(example_df)
```

---

# 📈 13. `head()` and `tail()`: Peek at Data

```{r}
# First few elements
head(numbers, 3)

# Last few elements
tail(numbers, 3)
```

---

# 📊 14. `table()`: Frequency Table

```{r}
# Frequency table of values
table(c("apple", "banana", "apple", "cherry"))
```

---

# ➕ 15. `paste()` and `paste0()`: Concatenate Strings

```{r}
# Paste with separator
paste("Hello", "world", sep = " ")

# Paste without separator
paste0("gene", 1:3)
```

---

# ⚖️ 16. `cut()`: Divide Numeric Data into Groups

```{r}
# Divide values into 3 groups
x <- c(1, 5, 10, 15, 20)
cut(x, breaks = 3)
```

---

# 🏛️ 17. `round()`, `floor()`, `ceiling()`: Rounding Functions

```{r}
val <- 3.75
round(val)    # Round to nearest
floor(val)    # Round down
ceiling(val)  # Round up
```

---

# 📐 18. `which()`: Get Index of Matching Condition

```{r}
# Index of elements > 5
which(numbers > 5)
```

---

# 💯 19. `any()` and `all()`: Logical Check Functions

```{r}
# Check if any or all elements match a condition
any(numbers > 5)  # TRUE if any value is > 5
all(numbers > 5)  # TRUE only if all values are > 5
```

---

# 🧪 20. `is.na()`: Check for Missing Values

```{r}
data <- c(1, 2, NA, 4)
is.na(data)       # Logical vector indicating NA
sum(is.na(data))  # Count of NAs
```

---

# 📉 21. `na.omit()`: Remove Missing Values

```{r}
# Remove missing values from a vector
data_no_na <- na.omit(data)
data_no_na
```

---

# 🧬 22. `length()`: Number of Elements

```{r}
# Get the number of elements in a vector
length(numbers)
```

---

# 🧩 23. `dim()`: Dimensions of Matrix or Data Frame

```{r}
m <- matrix(1:12, nrow = 3)
dim(m)
```

---

# 🛠️ 24. `names()`: Access or Set Names of Elements

```{r}
named_vec <- c(a = 1, b = 2, c = 3)
names(named_vec)
names(named_vec) <- c("x", "y", "z")
named_vec
```

---

# 📦 25. `append()`: Add Elements to a Vector

```{r}
vec <- c(1, 2, 3)
append(vec, 4)           # Add 4 at the end
append(vec, 0, after = 0) # Add 0 at the beginning
```

---

# ✅ Conclusion

These built-in R functions are essential tools for data manipulation and exploration. Get comfortable with them, and you'll have a strong foundation for more advanced work!
