---
title: "Introduction to Bioconductor in R"
author: "Chirag Parsania"
format: html
editor: visual
---

```{r setup, include=TRUE, message=FALSE,warning=F,}
# Install necessary Bioconductor packages if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Core Bioconductor packages
required_pkgs <- c(
  "GenomicRanges", "plyranges", "AnnotationDbi",
  "org.Hs.eg.db", "TxDb.Hsapiens.UCSC.hg38.knownGene", "ggplot2"
)

# Install missing packages
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg, ask = FALSE)
}

# Load packages
ll<-lapply(required_pkgs, library, character.only = TRUE, quietly = T)

library(GenomicRanges)
library(plyranges)
library(dplyr)
```

# 📖 Introduction

This tutorial introduces **Bioconductor**, a powerful ecosystem of R packages designed for the analysis and visualization of **high-throughput biological data**. While the audience is already familiar with **R**, this guide will focus on foundational Bioconductor concepts, especially how to represent genomic ranges and explore gene annotations using Bioconductor's tools.

# 📦 What is Bioconductor?

**Bioconductor** is an open-source project that develops and maintains R packages for the analysis of high-throughput biological data. Its main objective is to enable **reproducible**, **scalable**, and **interoperable** computational biology and bioinformatics workflows.

## 🔗 Project Information

-   **Website**: <https://bioconductor.org>
-   **Initial Release**: 2001
-   **Current Release**: Bioconductor 3.21
-   **Compatible R Version**: R 4.5.x
-   **Update Cycle**: Two major releases per year (aligned with R releases)
-   **Core Team**: Bioconductor Core Team at the Fred Hutchinson Cancer Center, Seattle
-   **Community**: Open-source contributors, package developers, and end users from around the globe

Bioconductor ensures that all packages meet strict **documentation**, **testing**, and **interoperability** standards. It is ideal for tasks such as:

-   RNA-seq and single-cell RNA-seq analysis
-   Genomic range operations (e.g., ChIP-seq, ATAC-seq)
-   Functional enrichment and pathway analysis
-   Integrative omics (genomics, transcriptomics, epigenomics)

## 🧠 Why Use Bioconductor?

- 🔄 **Designed for biology**: Handles genomic coordinates, gene IDs, metadata, etc.
- 🧩 **Interoperability**: Shared data classes and standards make package integration seamless.
- 🔒 **Reproducibility**: Versioned releases and standardized documentation.
- 🌍 **Community**: Large, supportive ecosystem of tutorials, forums, and conferences (e.g., BioC Asia, BioC Europe).


## 🔧 What Are Core Bioconductor Packages?

**Core Bioconductor packages** are foundational tools that provide standardized infrastructure and functionality used across many other packages. These include:

-   **`BiocGenerics`** – Generic functions shared across Bioconductor
-   **`S4Vectors`** – Provides S4 vector infrastructure
-   **`IRanges`** and **`GenomicRanges`** – Efficient handling of genomic intervals
-   **`AnnotationDbi`** – Framework for working with annotation databases
-   **`SummarizedExperiment`** – Container for assay + metadata integration

These core packages are maintained by the Bioconductor core team and serve as building blocks for downstream analysis workflows. They ensure consistency, interoperability, and performance across the Bioconductor ecosystem.

## 🧰 Bioconductor Package Ecosystem

Some of the most widely used categories include:

- 🧬 **RNA-seq**: `DESeq2`, `edgeR`, `limma`
- 🧠 **Functional Analysis**: `clusterProfiler`, `ReactomePA`, `topGO`
- 🧬 **Annotation Access**: `biomaRt`, `AnnotationHub`, `org.*`, `TxDb.*`
- 🧫 **Single-cell**: `Seurat`, `SingleCellExperiment`, `scran`
- 🧮 **Data Containers**: `SummarizedExperiment`, `SingleCellExperiment`
- 📊 **Visualization**: `ComplexHeatmap`, `ggbio`, `Gviz`




# 🧬 Representing Genomic Ranges with `GRanges` and `plyranges`

Bioconductor uses the `GRanges` class (from **GenomicRanges**) to represent genomic intervals. The **plyranges** package provides a **dplyr-like grammar** for working with these ranges.

```{r}



gr <- GRanges(
  seqnames = c("chr1", "chr1", "chr2"),
  ranges = IRanges(start = c(100, 150, 500), end = c(200, 400, 550)),
  strand = c("+", "-", "+")
)
gr

# Add a metadata column and mutate
gr <- as_granges(gr) %>%
  mutate(score = width) %>%
  arrange(dplyr::desc(score))


# Use plyranges to filter ranges
as_granges(gr) %>% 
  filter(score > 75)



# Reduce overlapping ranges
as_granges(gr) %>%
  reduce_ranges()

# Find overlaps with a new set
gr2 <- GRanges(
  seqnames = c("chr1", "chr2"),
  ranges = IRanges(c(180, 520), c(220, 560))
)

join_overlap_inner(as_granges(gr), as_granges(gr2))

```

# Visual Aid for GRanges joins

```         

# join_overlap_inner
gr1:   |======|         (100–200)
gr2:     |======|       (150–250)
result:  |====|         (150–200)


# join_overlap_left
gr1:    |======|         (100–200)
gr2:      |======|       (150–250)
result: |======|         (100–200)

# join_overlap_right
gr1:   |======|         (100–200)
gr2:     |======|       (150–250)
result:  |======|     (150–250)

# join_overlap_inner 

gr1:      |====| (100–200) 

gr2:         |====| (180–220)

Overlap:     |=| (180–200)


# join_overlap_inner (no overlap)
gr1:   |======|         (100–140)
gr2:             |======|   (160–200)
result: <empty>

# join_overlap_full
gr1:   |======|         (100–200)
gr2:     |======|       (150–250)
result: |======| +     (100–200)
        + |======|     (150–250)
```

# 🧭 Gene Annotation Using OrgDb

## What is an OrgDb?

An **OrgDb** is a type of annotation package in Bioconductor that provides curated gene-level annotations for a specific organism. These databases include mappings between various biological identifiers, such as gene symbols, Entrez IDs, Ensembl IDs, GO terms, chromosome locations, and gene product names.

The human OrgDb package is `org.Hs.eg.db`.

```{r}
library(org.Hs.eg.db)
```

You can explore available key types:

```{r}
keytypes(org.Hs.eg.db)
```

Common types include: - `SYMBOL`: gene symbols (e.g., "TP53") - `ENTREZID`: NCBI Entrez Gene IDs - `ENSEMBL`: Ensembl gene IDs - `GENENAME`: Full gene name

You can also see what columns (annotations) you can retrieve:

```{r}
columns(org.Hs.eg.db)
```

This includes: - `CHR`: chromosome - `GO`: Gene Ontology terms - `KEGG`: KEGG pathway IDs - `PFAM`: protein domains - `UNIPROT`: UniProt identifiers

## Example 1: Creating a Gene Annotation Table

```{r}
library(AnnotationDbi)

# Retrieve annotation for a set of gene symbols
genes <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = c("TP53", "EGFR", "BRCA1"),
  columns = c("SYMBOL", "GENENAME", "ENSEMBL", "ENTREZID", "CHR"),
  keytype = "SYMBOL"
)
genes

```

## Example 2: Retrieve GO Term Annotations

```{r}
go_annot <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = c("TP53", "EGFR", "BRCA1"),
  columns = c("SYMBOL", "GO", "ONTOLOGY"),
  keytype = "SYMBOL"
)
go_annot %>% head()
```

This table links gene symbols to Gene Ontology IDs and their categories (BP, MF, CC).

## Example 3: Retrieve KEGG Pathway Annotations

```{r}
kegg_annot <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = c("TP53", "EGFR", "BRCA1"),
  columns = c("SYMBOL", "PATH"),
  keytype = "SYMBOL"
)
kegg_annot %>% head()
```

This returns KEGG pathway identifiers associated with each gene.

# 🧱 Building Annotation Tables for Genes, Transcripts, Exons, Introns

```{r}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# Extract genes, transcripts, exons
gene_ranges <- genes(txdb)
transcript_ranges <- transcripts(txdb)
exon_ranges <- exons(txdb)

# Calculate introns from transcripts and exons
intron_ranges <- intronsByTranscript(txdb, use.names = TRUE)
```

# 📊 Plot Distribution of Gene Lengths

```{r}
library(ggplot2)

gene_df <- as.data.frame(gene_ranges)
gene_df$length <- gene_df$end - gene_df$start + 1

ggplot(gene_df, aes(x = length)) +
  geom_histogram(bins = 50, fill = "steelblue") +
  scale_x_log10() +
  labs(title = "Distribution of Gene Lengths", x = "Gene length (log scale)", y = "Count")
```

# 💡 Summary

In this first lecture, we: - Introduced Bioconductor and its purpose - Used `GRanges` and `plyranges` for genomic interval manipulation, including filtering, reducing, joining, and mutating ranges - Introduced `OrgDb` and thoroughly explored `org.Hs.eg.db`, its key types and annotation columns - Demonstrated how to retrieve GO and KEGG annotations using `AnnotationDbi::select()` - Built annotation tables for genes, transcripts, exons, and introns using a TxDb - Visualized the distribution of gene lengths

Next, we will explore **functional annotations**, such as GO terms and pathway mapping using clusterProfiler and biomaRt.

# 📚 Resources

-   <https://bioconductor.org>
-   [plyranges package](https://bioconductor.org/packages/release/bioc/html/plyranges.html)
-   [org.Hs.eg.db documentation](https://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html)
-   [TxDb.Hsapiens.UCSC.hg38.knownGene](https://bioconductor.org/packages/release/data/annotation/html/TxDb.Hsapiens.UCSC.hg38.knownGene.html)

------------------------------------------------------------------------

# ✨ Acknowledgements

Thanks to the R and Bioconductor communities for building and maintaining robust tools and documentation for genomics research.
