---
title: "Base R Practice Tutorial"
author: "Chirag Parsania - CI Bioinformatics Club"
format: html
editor: visual
---

# 📚 What You'll Practice

In this tutorial, you will practice the following foundational R concepts. Click a section to jump directly to it:

1.  [📏 1. Vectors and Basic Operations]
2.  [📦 2. Lists and Matrices]
3.  [📁 3. Data Frames and Factors]
4.  [🔍 4. Subsetting]
5.  [🔁 5. Loops and Conditionals]
6.  [🛠️ 6. Functions]
7.  [🧪 7. Practice Using Inbuilt R Functions]

Let’s jump into the practice!

# 📝 Practice Questions

## 📏 1. Vectors and Basic Operations

1.  Create a numeric vector of numbers from 1 to 10.

    ```{r}


    x <- c(1,2,3,4,5,7,8,9,10)

    x * 3

    fav_fruits  =  c("apple", "banana", "pineapple", "guava", "mango")

    ```

2.  Multiply each element of the vector by 3.

3.  Create a character vector of your favorite 5 fruits.

4.  Create a logical vector indicating whether elements of 1:10 are even.

    ```{r}

    numbers[numbers %% 2 != 0]



    ```

## 📦 2. Lists and Matrices

5.  Create a list containing your name, age, and a numeric vector of scores.
6.  Access the second score from the list.
7.  Create a 3x4 matrix of numbers from 1 to 12.
8.  Extract the element in the 3rd row and 2nd column of the matrix.

## 📁 3. Data Frames and Factors

9.  Create a data frame with columns: Name, Age, and Gender (as factor) for 4 people.
10. Add a column "Score" with random numbers between 60 and 100.
11. Filter the data frame to show only rows where Score \> 80.

## 🔍 4. Subsetting

12. From the data frame above, extract the "Name" and "Score" columns.
13. Select the first 2 rows of the data frame.
14. From a list of numbers, extract all values greater than 10.

## 🔁 5. Loops and Conditionals

15. Use a `for` loop to print numbers from 1 to 5.
16. Use a `while` loop to print numbers from 10 down to 6.
17. Use an `if-else` to print "Pass" if score \> 60, otherwise "Fail".

## 🛠️ 6. Functions

18. Write a function that takes a vector and returns its mean.
19. Write a function that checks whether a number is prime.
20. Write a function that counts how many numbers in a vector are greater than the mean.

## 🧪 7. Practice Using Inbuilt R Functions

21. Use `mean()` to calculate the average of a vector of your choice.
22. Create a character vector with repeated fruit names and use `table()` to get the frequency of each.
23. Create a vector with some missing values. Use `is.na()` and `na.omit()` to clean it.
24. Take a vector and:

-   Use `sort()` to sort it
-   Use `order()` to get the index order
-   Use `rev()` to reverse it

25. Write a custom mode function using `unique()` and `tabulate()` and apply it on a test vector.
26. Create a numeric vector of length 10 and generate 3 random samples using `sample()`.
27. Concatenate strings using `paste()` and `paste0()` to build variable names dynamically.
28. Generate a sequence of even numbers from 2 to 20 using `seq()`.
29. Check whether any or all values in a vector are greater than 5 using `any()` and `all()`.
30. Use `cut()` to divide a numeric vector into 4 bins.

------------------------------------------------------------------------

# ✅ Solutions

## 1. Vectors and Basic Operations

```{r}
vec <- 1:10
vec * 3
fruits <- c("Apple", "Banana", "Mango", "Pineapple", "Grape")
is_even <- vec %% 2 == 0
```

## 2. Lists and Matrices

```{r}
my_list <- list(name = "Alex", age = 28, scores = c(88, 76, 92))
my_list$scores[2]

my_matrix <- matrix(1:12, nrow = 3, ncol = 4)
my_matrix[3, 2]
```

## 3. Data Frames and Factors

```{r}
df <- data.frame(
  Name = c("Tom", "Sara", "Mike", "Lucy"),
  Age = c(23, 25, 22, 24),
  Gender = factor(c("M", "F", "M", "F"))
)

set.seed(123)
df$Score <- sample(60:100, 4)

df[df$Score > 80, ]
```

## 4. Subsetting

```{r}
df[, c("Name", "Score")]
df[1:2, ]

nums <- c(4, 9, 15, 20, 3)
nums[nums > 10]
```

## 5. Loops and Conditionals

```{r}
for (i in 1:5) {
  print(i)
}

x <- 10
while (x >= 6) {
  print(x)
  x <- x - 1
}

score <- 70
if (score > 60) {
  print("Pass")
} else {
  print("Fail")
}
```

## 6. Functions

```{r}
mean_func <- function(x) {
  mean(x)
}

is_prime <- function(n) {
  if (n < 2) return(FALSE)
  for (i in 2:sqrt(n)) {
    if (n %% i == 0) return(FALSE)
  }
  return(TRUE)
}

greater_than_mean <- function(x) {
  sum(x > mean(x))
}

# Test
mean_func(c(1, 2, 3, 4))
is_prime(11)
greater_than_mean(c(1, 5, 7, 2))
```

## 7. Practice Using Inbuilt R Functions

```{r}
# Q21: mean()
mean(c(4, 5, 6, 7, 8))

# Q22: table()
fruits <- c("apple", "banana", "apple", "mango", "banana", "apple")
table(fruits)

# Q23: is.na() and na.omit()
vec <- c(1, 2, NA, 4, NA, 6)
is.na(vec)
clean_vec <- na.omit(vec)
clean_vec

# Q24: sort(), order(), rev()
nums <- c(12, 3, 7, 1)
sort(nums)
order(nums)
rev(nums)

# Q25: custom mode function
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
getmode(c(1, 2, 2, 3, 3, 3, 4))

# Q26: sample()
x <- 1:10
sample(x, 3)

# Q27: paste() and paste0()
paste("gene", 1:5, sep = "_")
paste0("gene", 1:5)

# Q28: seq()
seq(2, 20, by = 2)

# Q29: any() and all()
y <- c(3, 7, 9)
any(y > 5)
all(y > 5)

# Q30: cut()
data <- c(5, 15, 25, 35, 45)
cut(data, breaks = 4)
```

# 🎉 Wrap-up

Great job working through these exercises! Keep practicing, explore more R functions, and don’t hesitate to experiment—because that’s where the real learning happens. Happy coding from the CI Bioinformatics Club! 🚀

------------------------------------------------------------------------
