---
title: "Base R Practice Tutorial"
author: "Chirag Parsania - CI Bioinformatics Club"
format: html
editor: visual
---

# 📝 Practice Questions

## 📏 1. Vectors and Basic Operations

1. Create a numeric vector of numbers from 1 to 10.
2. Multiply each element of the vector by 3.
3. Create a character vector of your favorite 5 fruits.
4. Create a logical vector indicating whether elements of 1:10 are even.

## 📦 2. Lists and Matrices

5. Create a list containing your name, age, and a numeric vector of scores.
6. Access the second score from the list.
7. Create a 3x4 matrix of numbers from 1 to 12.
8. Extract the element in the 3rd row and 2nd column of the matrix.

## 📁 3. Data Frames and Factors

9. Create a data frame with columns: Name, Age, and Gender (as factor) for 4 people.
10. Add a column "Score" with random numbers between 60 and 100.
11. Filter the data frame to show only rows where Score > 80.

## 🔍 4. Subsetting

12. From the data frame above, extract the "Name" and "Score" columns.
13. Select the first 2 rows of the data frame.
14. From a list of numbers, extract all values greater than 10.

## 🔁 5. Loops and Conditionals

15. Use a `for` loop to print numbers from 1 to 5.
16. Use a `while` loop to print numbers from 10 down to 6.
17. Use an `if-else` to print "Pass" if score > 60, otherwise "Fail".

## 🛠️ 6. Functions

18. Write a function that takes a vector and returns its mean.
19. Write a function that checks whether a number is prime.
20. Write a function that counts how many numbers in a vector are greater than the mean.

---

# ✅ Solutions

## 1. Vectors and Basic Operations

```{r}
vec <- 1:10
vec * 3
fruits <- c("Apple", "Banana", "Mango", "Pineapple", "Grape")
is_even <- vec %% 2 == 0
```

## 2. Lists and Matrices

```{r}
my_list <- list(name = "Alex", age = 28, scores = c(88, 76, 92))
my_list$scores[2]

my_matrix <- matrix(1:12, nrow = 3, ncol = 4)
my_matrix[3, 2]
```

## 3. Data Frames and Factors

```{r}
df <- data.frame(
  Name = c("Tom", "Sara", "Mike", "Lucy"),
  Age = c(23, 25, 22, 24),
  Gender = factor(c("M", "F", "M", "F"))
)

set.seed(123)
df$Score <- sample(60:100, 4)

df[df$Score > 80, ]
```

## 4. Subsetting

```{r}
df[, c("Name", "Score")]
df[1:2, ]

nums <- c(4, 9, 15, 20, 3)
nums[nums > 10]
```

## 5. Loops and Conditionals

```{r}
for (i in 1:5) {
  print(i)
}

x <- 10
while (x >= 6) {
  print(x)
  x <- x - 1
}

score <- 70
if (score > 60) {
  print("Pass")
} else {
  print("Fail")
}
```

## 6. Functions

```{r}
mean_func <- function(x) {
  mean(x)
}

is_prime <- function(n) {
  if (n < 2) return(FALSE)
  for (i in 2:sqrt(n)) {
    if (n %% i == 0) return(FALSE)
  }
  return(TRUE)
}

greater_than_mean <- function(x) {
  sum(x > mean(x))
}

# Test
mean_func(c(1, 2, 3, 4))
is_prime(11)
greater_than_mean(c(1, 5, 7, 2))
```

---
