---
title: "Introduction to Basic R Data Structures and Loops"
author: "Chirag Parsania"
date: "`r Sys.Date()`"
format: html
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# **Introduction**

R provides various **data structures** to store and manipulate data efficiently. In this guide, we will cover the basic data structures in R and explore **looping functions** such as `apply`, `sapply`, and `lapply`.

------------------------------------------------------------------------

## **1. Basic Data Structures in R**

### **a) Vectors**

A **vector** is the simplest data structure in R. It is a collection of elements of the same type. Vectors can be **numeric, character, or logical**. They are useful when dealing with **a series of values of the same kind**, such as numerical datasets or character lists.

```{r}
# Numeric vector 

num_vec <- c(1, 2, 3, 4, 5)
```

# Character vector

```{r}
char_vec <- c("apple", "banana", "cherry")
```

# Logical vector

```{r}

log_vec <- c(TRUE, FALSE, TRUE)
```

**Accessing Elements of a Vector:**

```{r}
num_vec[1]    # First element
char_vec[2]   # Second element
log_vec[3]    # Third element
num_vec[2:4]  # Elements from index 2 to 4
```

**Key Features:**

-   One-dimensional.

<!-- -->

-   Can store only one type of data.

<!-- -->

-   The most basic building block in R.

------------------------------------------------------------------------

### **b) Factors**

A **factor** is a categorical data structure used to store discrete values, such as labels or groups. It is particularly useful in statistical modeling and visualization.

```{r}
# Creating a factor variable
gender <- factor(c("Male", "Female", "Female", "Male", "Male"))
print(gender)
```

**Accessing Factor Levels:**

```{r}
levels(gender)  # Returns unique categories
table(gender)   # Counts occurrences of each category
gender[3]       # Access the third element
```

**Key Features:**

-   Stores categorical data efficiently.

-   Levels represent unique categories.

-   Useful in statistical modeling.

**Accessing Factor Levels:**

```{r}

levels(gender)  # Returns unique categories

table(gender)   # Counts occurrences of each category

```

------------------------------------------------------------------------

### **c) Matrices**

A **matrix** is a two-dimensional data structure where all elements are of the same type. Matrices are useful when working with **numerical data arranged in rows and columns**, such as correlation tables or mathematical operations.

```{r}
# Creating a 3x3 matrix
mat <- matrix(1:9, nrow = 3, byrow = TRUE)
print(mat)
```

**Accessing Rows and Columns:**

```{r}

# Accessing a specific row
mat[2, ]  # Returns second row

# Accessing a specific column
mat[, 3]  # Returns third column

# Accessing a specific element
mat[2, 3]  # Returns value at row 2, column 3


```

**Key Features:**

-   Two-dimensional (rows and columns).

-   Stores only one data type (numeric, character, or logical).

-   Useful for mathematical computations.

------------------------------------------------------------------------

### **d) Lists**

A **list** is a flexible data structure that can hold elements of **different types**, including vectors, matrices, and even other lists. It is useful for storing **heterogeneous data**, such as a mix of numbers, text, and tables.

```{r}
# Creating a list
my_list <- list(name = "John", age = 30, scores = c(90, 85, 88))
print(my_list)
```

**Key Features:**

-   Can store different data types.

-   Can contain other data structures like vectors, matrices, and even lists.

-   Elements are accessed using double square brackets (`[[ ]]`).

**Pros and Cons of Lists:**

| Pros | Cons |
|------------------------------------|------------------------------------|
| Can store multiple data types | More complex to manipulate compared to data frames |
| Flexible and allows nesting | Not as efficient for tabular data |
| Useful for returning multiple results in functions | Accessing elements requires `[[ ]]` notation |

------------------------------------------------------------------------

### **e) Data Frames**

A **data frame** is a table-like structure where each column can contain different types of data. It is the most commonly used structure for handling **tabular data** in R, such as survey responses or experimental data.

```{r}

# Creating a data frame
df <- data.frame(Name = c("Alice", "Bob", "Charlie"), 
                 Age = c(25, 30, 35), 
                 Score = c(90, 85, 88))
print(df)
```

**Accessing Rows and Columns:**

```{r}

# Accessing a specific column
df$Name  # Returns all names

# Accessing a specific row
df[2, ]  # Returns the second row

# Accessing a specific element
df[3, "Score"]  # Returns Score of the third row
```

**Key Features:**

-   Two-dimensional (rows and columns).

-   Each column can have a different data type.

-   Similar to a spreadsheet or SQL table.

-   Rows and columns are indexed for easy access.

**Pros and Cons of Data Frames:**

| Pros | Cons |
|------------------------------------|------------------------------------|
| Ideal for tabular data | Uses more memory compared to lists |
| Easy to manipulate with built-in functions | Requires uniform column lengths |
| Compatible with data analysis packages | Not suitable for storing complex objects |

------------------------------------------------------------------------

## **2. Looping in Base R**

### **a) Using `apply()` for Matrices**

The `apply()` function applies a function to **rows or columns of a matrix**.

```{r}

# Sum of each row
apply(mat, 1, sum)

# Sum of each column
apply(mat, 2, sum)
```

------------------------------------------------------------------------

### **b) Using `lapply()` for Lists**

The `lapply()` function applies a function to **each element of a list** and returns a list.

```{r}
# Applying sqrt() to each numeric element in a list
num_list <- list(a = 16, b = 25, c = 36)
lapply(num_list, sqrt)
```

------------------------------------------------------------------------

### **c) Using `sapply()` for Simpler Output**

The `sapply()` function is similar to `lapply()` but tries to simplify the output.

```{r}

sapply(num_list, sqrt)  # Returns a vector instead of a list
```

------------------------------------------------------------------------

### **d) Using `tapply()` for Grouped Calculations**

The `tapply()` function applies a function to **subsets of a vector** based on a factor.

```{r}
# Creating a sample dataset
scores <- c(85, 90, 78, 92, 88, 76)
group <- factor(c("A", "A", "B", "B", "A", "B"))

# Calculate mean score per group
tapply(scores, group, mean)
```

------------------------------------------------------------------------

## **3. Summary**

-   **Vectors, Matrices, Factors, Lists, and Data Frames** are essential data structures in R.
-   **Lists** can store different data types, while **data frames** are structured like tables.
-   **Factors** are useful for handling categorical data.
-   **Base R looping functions (`apply`, `sapply`, `lapply`, `tapply`)** help in data manipulation and analysis.

For more details, refer to the **official R documentation**: <https://cran.r-project.org/manuals.html>

------------------------------------------------------------------------

Happy Coding! 🚀
